---
title: "Change Point Detection"
output: 
  rmarkdown::html_document:
    theme: readable
vignette: >
  %\VignetteIndexEntry{Change Point Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A change point is an abrupt shift or a temporal pattern switch (e.g., beginning of a trend) in the time series.
In FFA, change points indicate inhomogeneous periods, meaning a single model may not represent the entire record adequately.
Instead, piecewise analysis should be applied to each homogeneous sub-period.
The FFA framework provides two statistical tests for change point detection: 

- The Pettitt test, for sudden changes in the mean.
- The Mann-Kendall-Sneyers (MKS) test, for detecting changes in trend.

This vignette will demonstrate how these statistical tests can be used together to robustly identify change points in the data.

## Case Study

This vignette will explore the Kootenai River at Porthill (08NH021) station, located on the border of British Columbia and Idaho.
The Porthill station is also downstream of the Libby Dam, which finished construction in 1972.
Data for this station is provided as `Application_2.csv` in the `ffaframework` package.

```{r, fig.width = 10, fig.height = 8, fig.align = "center", out.width = "100%"}
library(ffaframework)

csv_path <- system.file("extdata", "Application_2.csv", package = "ffaframework")
df <- read.csv(csv_path, comment.char = "#")
df <- subset(df, !is.na(max)) # Remove missing values

head(df)

plot_ams_data(df$max, df$year, title = "Kootenai River at Porthill (08NH021)")
```

## The Pettitt Test

This rank-based test detects a single abrupt change in the median of a time series.
The null hypothesis assumes no change point.

Use the `eda_pettitt_test` function to perform the test. It requires two arguments: 

- `data`: the annual maximum series (AMS)
- `years`: corresponding numeric vector of years

```{r, fig.width = 10, fig.height = 8, fig.align = "center", out.width = "100%"}
pettitt_test <- eda_pettitt_test(df$max, df$year)

print(pettitt_test$p_value)

print(pettitt_test$change_year)

plot_pettitt_test(pettitt_test)
```

**Conclusion**: A p-value of <0.001 provides strong evidence of a change point in the year 1972.

## The MKS Test

The Mann-Kendall-Sneyers (MKS) test identifies trend changes in the data.

Use `eda_mks_test` with the same arguments as above.

```{r, fig.width = 10, fig.height = 8, fig.align = "center", out.width = "100%"}
mks_test <- eda_mks_test(df$max, df$year)

print(mks_test$p_value)

print(mks_test$change_df$year)

plot_mks_test(mks_test)
```

**Conclusion**: At a p-value of 0.015, there is evidence of trend changes in 1960 and 1985.

**Note**: Since the MKS test can identify multiple change points, the reported p-value is determined using the *most significant* change point.

## Interpreting and Selecting Change Points

In this example, the Pettitt and MKS tests both suggest structural changes in the time series.

Consider the following guidelines when choosing where to split the data:

1. **Incorporate domain knowledge and case-specific understanding**. For example, if we knew that a water regulation structure was built in 1972 (the Libby dam, for this dataset), this would support the Pettitt test result.
2. **Avoid overpartitioning**. The Pettitt and MKS tests operate independently and may detect multiple change points. To reduce sample size issues and uncertainty, retain only significant change points with physical justification.
3. **Prioritize based on p-value**. Lower p-values indicate stronger evidence and should be given more weight.
