---
title: "Stationary FFA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stationary FFA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how to use the `ffaframework` package to perform flood frequency analysis (model selection, parameter estimation, uncertainty quantification, and model performance assessment) under the assumption of stationarity.

The results are expressed in terms of *return periods* and *return levels*.
The *return period* of a flood is the expected number of years between annual maximum events of equal or greater magnitude while the *return level* is the magnitude of the annual maximum event, measured in $\text{m}^3/\text{s}$.

## Case Study

This vignette will explore data from the Athabasca River at Athabasca (07BE001), an unregulated hydrological monitoring station.
A statistical analysis of the data from this station does not reveal any evidence of trends in the mean or variability of the data.
Data for this station is provided as `CAN-07BE001.csv` in the `ffaframework` package.

```{r, fig.width = 10, fig.height = 8, fig.align = "center", out.width = "100%"}
library(ffaframework)

df <- data_local("CAN-07BE001.csv")
head(df)

plot_ams_data(df$max, df$year, title = "Athabasca River at Athabasca (07BE001)")
```

## Distribution Selection

First, a suitable probability distribution for the data is selected using the L-moments.

1. `select_ldistance` chooses the distribution with theoretical L-skewness ($\tau_3$) and L-kurtosis ($\tau_4$) closest to the sample, based on Euclidean distance.
2. `select_lkurtosis` selects the distribution with the smallest difference between sample and theoretical L-kurtosis (for three-parameter distributions only).
3. `select_zstatistic` uses a fitted 4-parameter Kappa distribution to estimate the sampling distribution of the L-kurtosis and selects the distribution with the smallest z-statistic.

```{r, fig.width = 10, fig.height = 8, fig.align = "center", out.width = "100%"}
selection <- select_ldistance(df$max)

print(selection$recommendation)

plot_lmom_diagram(selection)
```

**Recommendation**: Use the generalized extreme value (GEV) distribution.

**Note**: For information about the other distributions, see the `selection$metrics` item.

You can find more information about the probability distributions supported by the framework [here](https://rileywheadon.github.io/ffa-docs/probability-distributions/).

## Parameter Estimation

The `ffaframework` package provides three methods for parameter estimation.
See [here](https://rileywheadon.github.io/ffa-docs/parameter-estimation/) for more information.

- `fit_lmoments`: L-moments parameter estimation.
- `fit_mle`: Maximum likelihood parameter estimation. 
- `fit_gmle`: Generalized maximum likelihood parameter estimation.

```{r, fig.width = 10, fig.height = 8, fig.align = "center", out.width = "100%"}
fit <- fit_lmoments(df$max, "GEV")

print(fit$params)

plot_sffa_fit(fit)
```

**Conclusion**: The GEV distribution with `location = 1600`, `scale = 616`, and `shape = 0.12` will be used.

## Uncertainty Quantification

Once a probability distribution is fitted, return levels (design events) can be readily estimated.
However, point estimates alone can be misleading --it is more informative to report confidence intervals to reflect estimation uncertainty.
The `uncertainty_bootstrap` function performs uncertainty quantification using the parametric bootstrap method. It requires three arguments:

- `data`: A vector of annual maximum series data.
- `model`: A three-letter code for a probability distribution (ex. `"GEV"`).
- `method`: A parameter estimation method. Must be `"L-moments"`, `"MLE"`, or `"GMLE"`.

By default, return levels are computed 2-, 5-, 10-, 20-, 50-, and 100- year return periods. 

```{r, fig.width = 10, fig.height = 8, fig.align = "center", out.width = "100%"}
uncertainty <- uncertainty_bootstrap(df$max, "GEV", "L-moments")

plot_sffa_estimates(fit, ci = uncertainty$ci)
```

**Example Conclusion**: Every 10 years, we can expect a flood of roughly $3{\small,}200\text{m}^3/\text{s}$ or greater.

## Model Assessment

Model performance is assessed using `model_assessment`, which reports a collection of assessment statistics about the flood frequency analysis.
`plot_model_assessment` compares the empirical plotting positions ("Observed Quantiles") and the predictions of the parametric model ("Model Quantiles"). 
The black line represents a perfect 1:1 correspondence between the model and the data.

```{r, fig.width = 10, fig.height = 8, fig.align = "center", out.width = "100%"}
assessment <- model_assessment(df$max, "GEV", fit$params, ci = uncertainty$ci)

plot_model_assessment(assessment)
```

**Conclusion**: The parametric model generally matches the plotting positions.
There is a small positive bias from $3{\small,}500\text{m}^3/\text{s}$ to $5{\small,}000\text{m}^3/\text{s}$.
